%\documentclass[11pt,twoside]{nsf_jwg} %!PN
\documentclass[letterpaper]{article}
\usepackage{amssymb}
\usepackage[cm]{fullpage}
\usepackage{amsmath}
\usepackage{epsfig,float,alltt}
\usepackage{psfrag,xr}
\usepackage[T1]{fontenc}
\usepackage{url}
\usepackage{pdfpages}
%\includepdfset{pagecommand=\thispagestyle{fancy}}

%
%***********************************************************************
%               New Commands
%***********************************************************************
%
%
\newcommand{\rb}[1]{\raisebox{1.5ex}{#1}}
 \newcommand{\trace}{\mathrm{trace}}
\newcommand{\real}{\mathbb R}  % real numbers  {I\!\!R}
\newcommand{\nat}{\mathbb R}   % Natural numbers {I\!\!N}
\newcommand{\cp}{\mathbb C}    % complex numbers  {I\!\!\!\!C}
\newcommand{\pp}{\mathbb P}      %{I\!\!\!\!P}
\newcommand{\ds}{\displaystyle}
\newcommand{\mf}[2]{\frac{\ds #1}{\ds #2}}
\newcommand{\Luenberger}[2]{{Luenberger, Page~#1, }{Prob.~#2}}
\newcommand{\Nagy}[2]{{Nagy, Page~#1, }{Prob.~#2}}
\newcommand{\spanof}[1]{\textrm{span} \{ #1 \}}
 \newcommand{\cov}{\mathrm{cov}}
 \newcommand{\E}{\mathcal{E}}
 \newcommand{\Expectof}[1]{{\cal E} \{ #1 \}}
  \newcommand{\ExpectofGiven}[2]{{\cal E} \{ #1 | #2 \}}
  \newcommand{\Covof}[2]{ \mathrm{cov} \left(#1,#2\right)}
\parindent 0pt

\newcommand{\bline}[1]{\underline{\hspace*{#1}}}
%
%
%***********************************************************************
%
%               End of New Commands
%
%***********************************************************************
%
%

\begin{document}

\vspace*{.1in}
\begin{center}
\LARGE \bf
ROB 501 Exam-I \\
\end{center}

\vspace*{1cm}

\subsection*{Time, Place, Rules and Review Session}

\begin{itemize}
\item The exam is Tuesday November 3, 6:10 PM to 8 PM. I will start passing out exams at 6 PM.
\item The exam rooms are assigned by First Letter of Last Name: \textbf{EECS 1311 (A-K) and EECS 1303 (L-Z)}
\item If you have a conflict with the exam time, please contact me no later than Tuesday October 27, 4 PM. Also, in your email, please include the nature of your conflict.
\item Rules are exactly those on the first page of Exam I 2014. Note that you cannot use your cell phone as a clock.
\item Review Session:  Sunday, Nov. 1, 5:10 PM to 6:30 PM in room 1311 EECS. This will be a Q\&A session. I am NOT lecturing. I will do my very best to answer questions that you pose. If there are no questions, we go home! Please bring questions.
    \item We will use at least 50\% of Tuesday's lecture (day of exam) also for Q\&A. I'll let you know ahead of time if I am lecturing the other 50\% or not. I do know yet.
\end{itemize}

\subsection*{Material Covered}

\begin{itemize}
\item From Lecture 1 through 22 October.
\item HW 01 through HW 06, except:
\begin{itemize}
\item No Lagrange multipliers
\item Nothing on probability
\item No weighted least squares
\item No questions on forgetting factors (because that is weighted least squares!)
\item No questions on Schur Complements. While positive definite matrices \underline{may be included} on the exam, you will not need to use Schur complements for any question on positive definite matrices.
\end{itemize}
\item RLS is covered, but once again, do not worry about forgetting factors.
\item You may have to invert by hand a $2 \times 2$ matrix.
\end{itemize}

\subsection*{Type of Questions}

\begin{itemize}

\item Your exam looks exactly like Exam 1 from 2014; see the CTOOLS site.

\item Regarding Exam 2 (in class) from 2014, ONLY question 6 on RLS is relevant to you. The other problems involve least squares for under determined systems of equations, Schur Complements, normal (Gaussian) random variables, QR factorizations, BLUE and MVE estimators. None of the latter topics are on the exam!

    \item Nothing on 2014 Exam 2 (take home) is relevant to your exam! Nothing. Zero.

\item I do not have any practice problems other than the 2014 Exams on the CTOOLS site. If I had other problems, I would gladly give them to you. Last year's class did not even have the old exams to look at!

    \item There will be ONE proof to give on the exam. When doing a proof, you can use as a fact ANYTHING we have established in lecture or HW.

    \item Because I have to grade 60+ exams this year, instead of the 30 I had last year, do NOT be surprised if the proof can be given in just a few lines. I am not promising that the proof will be very short, but if you happen to find a very short proof, do not get all worried! Also, any proof I ask you to give can be worked in multiple ways. If your way is correct, you will get full credit even if it is not the method I used in the solutions.

        \item If you give more than one proof or solution to a problem, you must tell me which one to grade. If you do not tell me which one to grade, I will grade the first one, even if it is wrong and something later is correct. What else can I do? The only reason I mention this is because it has come up in the past.

        \item Everyone always wants to come to me and ask if they have shown enough on the workout problems or the proofs. \textbf{I cannot answer that question.} My best advice is to show your work clearly. Show the steps you are following. You do NOT need to re-derive something we have established in class or HW. You can just state it as a fact and then use it.

\subsection*{Suggested Strategy}

Spend at most 25 minutes initially on the multiple choice questions. Mark the ones you are sure of and move on to the work-out problems. Then come back to the multiple choice questions at the end. Yeah, I know, we all hate multiple choice questions, but it is the only way to have a broad coverage of the material with very few calculations. When you look at them, you will see that our multiple choice questions actually consist of four T/F questions, worth 2 points each. You circle only the responses that are true, and it is never the case that all responses are true nor all are false for a given question. Note also that if you mark all as true (or all as false), then you get no credit whatsoever because it is assumed that you are just guessing. So, not matter what, even if you are guessing, please do mark at least ONE as T and DO NOT MARK ALL as T.

%Last year, the first person done with the exam completed it in 35 minutes, with a score of 78 out of 80 ... and it was a first-year student.

\end{itemize}


\end{document} 